---
title: "Fig S7"
output:
  md_document: default
  word_document: default
  pdf_document: default
  html_notebook: default
---

#### Background
This is code corresponding to Fig S7. Following a reviewer's suggestion, we assessed if our results were robust to the definition of highly expressed genes. Here, we use CUB calculated after defining HEG as genes with lowest 1% ENC' in every genome.

#### Setup
Read and process the data required for this analysis-
```{r read_data }
library(ape)

full_tree_oi <- read.tree(file = "trees/bacteria_pruned_genome_ids.newick")
tips_oi <- read.csv(file = "genome_lists/Gammaproteo_curated_genomes_list.txt", stringsAsFactors = F, strip.white = T, header = F)
tree_oi <- drop.tip(phy = full_tree_oi, tip = setdiff(full_tree_oi$tip.label, tips_oi$V1))

rrnCN_rrnDB <- read.table(file="traits_data/rRNA_CN.txt", header = T, row.names = 1, stringsAsFactors = F)
full_aaUsage_HEG <- read.table(file ="traits_data/aaUsage_HEG_p0.01.txt", header = T, row.names=1, stringsAsFactors = F)
full_GCN_data <- read.table(file ="traits_data/tRNA_GCN.txt", header = T, row.names=1, stringsAsFactors = F)
full_aaENCp_HEG <- read.table(file ="traits_data/ENCp_HEG_aawise_p0.01.txt", header = T, row.names=1, stringsAsFactors = F)
full_aaENCp_rest <- read.table(file ="traits_data/ENCp_rest_aawise_p0.01.txt", header = T, row.names=1, stringsAsFactors = F)
full_aaS <- read.table(file ="traits_data/S_aawise_p0.01.txt", header = T, row.names=1, stringsAsFactors = F)

rrnaCN2 <- rrnCN_rrnDB[tree_oi$tip.label, "rRNA_CN_final"]
rrnaCN2 <- round(rrnaCN2, digits = 0)
names(rrnaCN2) <- tree_oi$tip.label
aaUsage_HEG <- full_aaUsage_HEG[tree_oi$tip.label,]
aaENCp_HEG <- full_aaENCp_HEG[tree_oi$tip.label, ]
aaENCp_rest <- full_aaENCp_rest[tree_oi$tip.label, ]
aaS <- full_aaS[tree_oi$tip.label, ]
  
GCN_data <- full_GCN_data[tree_oi$tip.label,]
tRNAs_oi <-setdiff(colnames(GCN_data), c("TGC", "GAT")) #ignore tRNAs associated with rRNA operons
tRNAtot <- rowSums(x = GCN_data[tRNAs_oi]) #total tRNA genes per genome

codon_table <- read.csv(file = "codon_aa_map.txt", header = T, stringsAsFactors=F)
aa_unique <- setdiff(unique(codon_table$aa), c("Stop", "Met", "Trp"))
redundancy_oi <- sapply(aa_unique, function(x) codon_table[which(codon_table$aa==x), "degen"])
```

#### deltaENCp and rRNA CN
```{r ENCp_slopes }
redundancy_mat <- matrix(data = rep(x = c(2,2,2,2,2,2,2,2,2,2,3,4,4,4,4,4,4,6,6), times = nrow(aaENCp_HEG)), nrow = nrow(aaENCp_HEG), ncol = length(aa_unique), byrow = T)
aaENCp_delta <- (aaENCp_rest - aaENCp_HEG)/(redundancy_mat-1)

intercepts_deltaENC_aa <- c()
intercepts_deltaENC_aa_sd <- c()
slopes_deltaENC_aa <- c()
slopes_deltaENC_aa_sd <- c()

par(mfcol=c(5,4), mar=c(2,2,0,1), oma=c(3,4,5,3))

for(aa_oi in aa_unique){
    if(redundancy_oi[[aa_oi]][1]<4){ylim1 <- -0.6} else {ylim1 <- -0.3}
    if(redundancy_oi[[aa_oi]][1]<4){ylim2 <- 1} else {ylim2 <- 0.6}
    
    plot(jitter(rrnaCN2,amount=0.15), aaENCp_delta[,aa_oi], cex=0.3, xlab="", ylab="", col=grey(0.5), xaxt="n", ylim=c(ylim1,ylim2), cex.axis=1) #
    loess_aa <- loess(aaENCp_delta[,aa_oi]~rrnaCN2, na.action=na.omit, span = 0.9)
    x_order <- order(loess_aa$x)
    lines(loess_aa$x[x_order], loess_aa$fitted[x_order], col="black", lwd=1)
    abline(a=0,b=0,col=grey(0.5),lty=2)
    
    cor_deltaENCp_rrnaCN2 <- cor.test(x = rrnaCN2, y = aaENCp_delta[,aa_oi], alternative = "greater", method = "spearman")
    pval_deltaENCp_rrnaCN2 <- format(cor_deltaENCp_rrnaCN2$p.val, digits = 2, scientific = T)
    rho_deltaENCp_rrnaCN2 <- format(cor_deltaENCp_rrnaCN2$estimate, digits = 2)
    legend("bottomright", legend = bquote(rho ~ "=" ~ .(rho_deltaENCp_rrnaCN2)) , bty="n", cex=1)
    legend("topright", legend = paste0("p = ", pval_deltaENCp_rrnaCN2), bty="n", cex=1)

    if(cor_deltaENCp_rrnaCN2$p.val<1e-2) {sigCol <- "black"} else {sigCol <- "coral"}
    legend("topleft", legend = aa_oi, bty="n", cex = 1.2, text.col = sigCol)
    
    lm_aa <- lm(aaENCp_delta[,aa_oi]~rrnaCN2)
    intercept_oi <- lm_aa$coefficients[1] + lm_aa$coefficients[2] # value at rRNA CN = 1
    intercepts_deltaENC_aa <- c(intercepts_deltaENC_aa, round(intercept_oi, digits = 3))
    slopes_deltaENC_aa <- c(slopes_deltaENC_aa, round(lm_aa$coefficients[2], digits = 3))
    summ_lm_aa <- summary(lm_aa)
    intercepts_deltaENC_aa_sd <- c(intercepts_deltaENC_aa_sd, round(summ_lm_aa$coefficients[1,2], digits = 3))
    slopes_deltaENC_aa_sd <- c(slopes_deltaENC_aa_sd, round(summ_lm_aa$coefficients[2,2], digits = 3))
}

names(intercepts_deltaENC_aa) <- names(intercepts_deltaENC_aa_sd) <- names(slopes_deltaENC_aa) <- names(slopes_deltaENC_aa_sd) <- aa_unique

plot(rrnaCN2, aaENCp_delta[,aa_oi], xlab="", ylab="", ylim=c(-0.3,0.6), col=grey(0.5), cex=0, cex.axis=1, xat=c(2,6,10,14))
axis(side = 4, at = c(0, 0.4), cex = 0.8)
legend("topleft", legend = "loess fit", bty="n", lty=1)

title(main="", xlab="rRNA CN", ylab=expression(paste(Delta, "ENC'")),outer=T, line=1, cex.lab=1.5)
```
#### Amino acid specific S and rRNA CN
```{r S_slopes }
intercepts_S_aa <- c()
intercepts_S_aa_sd <- c()
slopes_S_aa <- c()
slopes_S_aa_sd <- c()

par(mfcol=c(4,3), mar=c(2,2,0,1), oma=c(3,4,2,3))

degen2_aas <- unique(codon_table[codon_table$degen==2, "aa"])
aas_oi <- c(degen2_aas, "Ile")

for(aa_oi in aas_oi){
    valid_aaS <- which(is.finite(aaS[,aa_oi]))
    aaS_oi <- abs(aaS[valid_aaS,aa_oi])

    ylim1 <- max(min(aa_oi, na.rm = T))
    ylim2 <- max(aaS_oi, na.rm = T)
    plot(jitter(rrnaCN2[valid_aaS],amount=0.15), aaS_oi, cex=0.3, xlab="", ylab="", col=grey(0.5), xaxt="n", cex.axis=1.2, ylim=c(0,2.5), cex.axis=1) #
    loess_aa <- loess(aaS_oi~rrnaCN2[valid_aaS], na.action=na.omit)
    x_order <- order(loess_aa$x)
    lines(loess_aa$x[x_order], loess_aa$fitted[x_order], col="black", lwd=1, span=0.9)

    cor_aaS_rrnaCN2 <- cor.test(x = rrnaCN2[valid_aaS], y = aaS_oi, alternative = "greater", method = "spearman")
    pval_aaS_rrnaCN2 <- format(cor_aaS_rrnaCN2$p.val, digits = 2, scientific = T)
    rho_aaS_rrnaCN2 <- format(cor_aaS_rrnaCN2$estimate, digits = 2)
    legend("bottomright", legend = bquote(rho ~ "=" ~ .(rho_aaS_rrnaCN2)) , bty="n", cex=1)
    legend("topright", legend = paste0("p = ", pval_aaS_rrnaCN2), bty="n", cex=1)

    if(cor_aaS_rrnaCN2$p.val<1e-2) {sigCol <- "black"} else {sigCol <- "coral"}
    legend("topleft", legend = aa_oi, bty="n", cex = 1.2, text.col = sigCol)
    
    lm_aa <- lm(aaS_oi[valid_aaS]~rrnaCN2[valid_aaS])
    intercept_oi <- lm_aa$coefficients[1] + lm_aa$coefficients[2] # value at rRNA CN = 1
    intercepts_S_aa <- c(intercepts_S_aa, round(intercept_oi, digits = 3))
    slopes_S_aa <- c(slopes_S_aa, round(lm_aa$coefficients[2], digits = 3))
    summ_lm_aa <- summary(lm_aa)
    intercepts_S_aa_sd <- c(intercepts_S_aa_sd, round(summ_lm_aa$coefficients[1,2], digits = 3))
    slopes_S_aa_sd <- c(slopes_S_aa_sd, round(summ_lm_aa$coefficients[2,2], digits = 3))
}
names(intercepts_S_aa) <- names(intercepts_S_aa_sd) <- names(slopes_S_aa) <- names(slopes_S_aa_sd) <- c(degen2_aas, "Ile")

plot(rrnaCN2[valid_aaS], aaS_oi, xlab="", ylab="", ylim=c(-2,4), col=grey(0.5), yaxt="n", cex=0, cex.axis=1, at=c(2,6,10,14))
legend("topleft", legend = "loess fit", bty="n", lty=1)

title(main="", xlab="rRNA CN", ylab= expression("S"[CUB]), outer=T, line=1, cex.lab=1.3)
```
#### Figure S7
CUB and amino acid usage, and tRNA GCN vs rRNA CN slopes
```{r FigS7, fig.width=7, fig.height=7 }
# svg(filename = "figures/Mahajan_FigS7.svg", width=7, height=7)
par(mfrow=c(2,2), mar=c(5,5,1,1))

median_aa_usage <- sapply(aa_unique, function(x) median(aaUsage_HEG[,x]))
lq_aa_usage <- sapply(aa_unique, function(x) quantile(aaUsage_HEG[,x], probs = 0.25))
hq_aa_usage <- sapply(aa_unique, function(x) quantile(aaUsage_HEG[,x], probs = 0.75))
single_rrnaCN <- which(rrnaCN2==1)
median_aa_basal <- sapply(aa_unique, function(x) median(aaUsage_HEG[single_rrnaCN,x]))
lq_aa_basal <- sapply(aa_unique, function(x) quantile(aaUsage_HEG[single_rrnaCN,x], probs = 0.25))
hq_aa_basal <- sapply(aa_unique, function(x) quantile(aaUsage_HEG[single_rrnaCN,x], probs = 0.75))

plot(median_aa_usage, slopes_deltaENC_aa, xlab="median amino acid usage in HEGs", ylab=expression(paste("slope of ", Delta, "ENC' vs rRNA CN")), pch=19, xlim=c(0,0.12), ylim=c(-0.03,0.08), col=col_oi, cex.axis=1.1, cex.lab=1.1)
text(median_aa_usage, slopes_deltaENC_aa, labels = names(median_aa_usage), pos=4, cex=0.9)
cor_slopes <- cor.test(x = median_aa_usage, y = slopes_deltaENC_aa, method = "spearman", alternative = "less")
rho_oi <- format(cor_slopes$estimate, digits = 2)
p_oi <- format(cor_slopes$p.value, digits = 2)
legend("topleft", legend = bquote(rho ~ " = " ~ .(rho_oi)), bty="n", cex=1.1)
legend("topright", legend = c(paste0("p = ", p_oi)), bty="n", cex=1.1)
arrows(median_aa_usage, slopes_deltaENC_aa-slopes_deltaENC_aa_sd, median_aa_usage, slopes_deltaENC_aa+slopes_deltaENC_aa_sd, length=0.0, angle=90, code=3, col=col_oi)
arrows(lq_aa_usage, slopes_deltaENC_aa, hq_aa_usage, slopes_deltaENC_aa, length=0.0, angle=90, code=3, col=col_oi)

plot(median_aa_usage[c(degen2_aas, "Ile")], slopes_S_aa, xlab="median amino acid usage in HEGs", ylab=expression(paste("slope of S"[CUB], " vs rRNA CN")), pch=19, col="coral", xlim=c(0,0.12), ylim=c(-0.05,0.15), cex.axis=1.1, cex.lab=1.1)
text(median_aa_usage[c(degen2_aas, "Ile")], slopes_S_aa, labels = names(median_aa_usage[c(degen2_aas, "Ile")]), pos=4, cex=0.9)
cor_slopes <- cor.test(x = median_aa_usage[c(degen2_aas, "Ile")], y = slopes_S_aa, method = "spearman", alternative = "less")
rho_oi <- format(cor_slopes$estimate, digits = 2)
p_oi <- format(cor_slopes$p.value, digits = 2)
legend("topleft", legend = bquote(rho ~ " = " ~ .(rho_oi)), bty="n", cex=1.1, text.col = "coral")
legend("topright", legend = c(paste0("p = ", p_oi)), bty="n", cex=1.1, text.col = "coral")
arrows(median_aa_usage[c(degen2_aas, "Ile")], slopes_S_aa-slopes_S_aa_sd, median_aa_usage[c(degen2_aas, "Ile")], slopes_S_aa+slopes_S_aa_sd, length=0.0, angle=90, code=3, col="coral")
arrows(lq_aa_usage[1:11], slopes_S_aa, hq_aa_usage[1:11], slopes_S_aa, length=0.0, angle=90, code=3, col="coral")

plot(slopes_tRNA_GCN_aa, slopes_deltaENC_aa, xlab="slope of tRNA GCN vs rRNA CN", ylab=expression(paste("slope of ", Delta, "ENC' vs rRNA CN")), pch=19, col=col_oi, xlim=c(0,0.75), ylim=c(-0.03,0.08), cex.axis=1.1, cex.lab=1.1)
text(slopes_tRNA_GCN_aa, slopes_deltaENC_aa, labels = names(slopes_tRNA_GCN_aa), pos=4, cex=0.9)
cor_slopes <- cor.test(x = slopes_tRNA_GCN_aa, y = slopes_deltaENC_aa, method = "spearman", alternative = "less")
rho_oi <- format(cor_slopes$estimate, digits = 2)
p_oi <- format(cor_slopes$p.value, digits = 2)
legend("topleft", legend = bquote(rho ~ " = " ~ .(rho_oi)), bty="n", cex=1.1)
legend("topright", legend = c(paste0("p = ", p_oi)), bty="n", cex=1.1)
arrows(slopes_tRNA_GCN_aa-slopes_tRNA_GCN_aa_sd, slopes_deltaENC_aa, slopes_tRNA_GCN_aa+slopes_tRNA_GCN_aa_sd, slopes_deltaENC_aa, length=0.0, angle=90, code=3, col=col_oi)
arrows(slopes_tRNA_GCN_aa, slopes_deltaENC_aa-slopes_deltaENC_aa_sd, slopes_tRNA_GCN_aa, slopes_deltaENC_aa+slopes_deltaENC_aa_sd, length=0.0, angle=90, code=3, col=col_oi)
    
plot(slopes_tRNA_GCN_aa[c(degen2_aas, "Ile")], slopes_S_aa, xlab="slope of tRNA GCN vs rRNA CN", ylab=expression(paste("slope of ", S[CUB], " vs rRNA CN")), pch=19, col=col_oi, xlim=c(0,0.75), ylim=c(-0.05, 0.15), cex.axis=1.1, cex.lab=1.1)
text(slopes_tRNA_GCN_aa[c(degen2_aas, "Ile")], slopes_S_aa, labels = names(slopes_tRNA_GCN_aa[c(degen2_aas, "Ile")]), pos=4, cex=0.9)
cor_slopes <- cor.test(x = slopes_tRNA_GCN_aa[c(degen2_aas, "Ile")], y = slopes_S_aa, method = "spearman", alternative = "less")
rho_oi <- format(cor_slopes$estimate, digits = 2)
p_oi <- format(cor_slopes$p.value, digits = 2)
legend("topleft", legend = bquote(rho ~ " = " ~ .(rho_oi)), bty="n", cex=1.1, text.col = "coral")
legend("topright", legend = c(paste0("p = ", p_oi)), bty="n", cex=1.1, text.col = "coral")
arrows(slopes_tRNA_GCN_aa[c(degen2_aas, "Ile")]-slopes_tRNA_GCN_aa_sd[c(degen2_aas, "Ile")], slopes_S_aa, slopes_tRNA_GCN_aa[c(degen2_aas, "Ile")]+slopes_tRNA_GCN_aa_sd[c(degen2_aas, "Ile")], slopes_S_aa, length=0.0, angle=90, code=3, col=col_oi)
arrows(slopes_tRNA_GCN_aa[c(degen2_aas, "Ile")], slopes_S_aa-slopes_S_aa_sd, slopes_tRNA_GCN_aa[c(degen2_aas, "Ile")], slopes_S_aa+slopes_S_aa_sd, length=0.0, angle=90, code=3, col=col_oi)
# dev.off()
```